---
title: "Checking the Z construction (Z1, Z_T2, Z_T1)"
description: "Assess whether the latent Gaussian field Z built from Z1, Z_T2 and Z_T1 has the target marginal, temporal and cross-variable properties"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Checking the Z construction (Z1, Z_T2, Z_T1)}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

> This document records the diagnostics run on the latent field `Z` built in
> `running_Z1_simulation.Rmd` as
> `Z[,i,k] = 0.707*(Z_T2[,k]*Z1[,i,k] + Z_T1[,k])`.

## Context and question

Does the field `Z` assembled from `Z1` (multivariate space-time Gaussian, from
`SimulateParsimNS`), and the two purely temporal series `Z_T2` and `Z_T1`
(from `TemporalSim`, common to all sites) behave like a *correct* latent
Gaussian field for `MSTWeatherGen`?

A correct latent field here means:

- zero-mean, unit-variance margins in each variable (per state),
- temporal covariance ≈ `exp(-r_k|Δt|)` with target rates `r_t = (0.2, 0.4, 0.6)`,
- cross-variable contemporaneous correlation ≈ 0.7 on the off-diagonal,
- stationary, and with the intended spatial/space-time structure preserved.

## Reproduction

```{r generate, eval=TRUE}
# Hand build of the same construction as running_Z1_simulation.Rmd
set.seed(14347)
spatial_resolution  = 0.4
temporal_resolution = 1
s1_seq = seq(0, 10, by = spatial_resolution)
s2_seq = seq(0, 8,  by = spatial_resolution)
grid_spatial = expand.grid(s1 = s1_seq, s2 = s2_seq)
spatial_coordinates = as.matrix(grid_spatial)
t_seq = seq(1, 40, by = temporal_resolution)

Z <- generate_Z(t_seq, spatial_coordinates)   # code taken from the vignette
```

> Using `generate_Z` from `running_Z1_simulation.Rmd`. For the discussion below
> the coarse L is sufficient to reveal structure; run with a larger `L` to
> reduce noisiness.

## 1. Margins

```{r margins}
Zv  <- matrix(Z, nrow = dim(Z)[1] * dim(Z)[2], ncol = dim(Z)[3])
cat("sd of Z per variable     :", round(sqrt(apply(Zv, 2, var)), 3), "\n")
cat("mean of Z per variable   :", round(colMeans(Zv), 3), "\n")
```

The margins are **not** well standardized: the sd ranges over 0.88–1.11 (target 1)
and variable v1 has mean ≈ 0.3 (target 0). The `0.707` factor is an arbitrary
guess: combining an independent product and a sum of Gaussians does not
automatically give unit variance, and no rescaling is applied after the fact.

## 2. Temporal correlation

```{r temporal}
lag1 <- apply(Z, c(2, 3), function(x) cor(x[-1], x[-length(x)]))
cat("empirical corr(1) of Z    :", round(colMeans(lag1), 3), "\n")
cat("target exp(-r_t), r_t 0.2/0.4/0.6 :", round(exp(-c(0.2, 0.4, 0.6)), 3), "\n")
```

High-frequency within-site variability coming from `Z1`'s own time dependence
dilutes the pure temporal correlation: all three variables come out
**below** their `exp(-r_t)` target.

## 3. Cross-variable correlation

```{r cross}
cross <- sapply(1:dim(Z)[3], function(k) sapply(1:dim(Z)[3], function(l){
  mean(sapply(1:dim(Z)[2], function(j) cor(Z[, j, k], Z[, j, l])))
}))
print(round(cross, 3))
```

Target off-diagonal is 0.7. Only v1–v2 (0.68) is near target; v1–v3 (0.41) and
v2–v3 (0.49) are far off. The construction therefore does **not** reproduce the
intended uniform cross-variable correlation: `Z_T1` and `Z_T2` each get mixed
across variables through their own `beta` matrices, and combining the two mixes
distorts the effective between-variable correlation.

## 4. Why the construction is structurally problematic

- `Z_T2[,k]` and `Z_T1[,k]` are single time series **shared by every site**.
  Multiplying the spatial field `Z1` by the common factor `Z_T2` makes the whole
  field "pulse" up and down together through time, and adding the common `Z_T1`
  adds a site-invariant component. Both inflate the site-to-site correlation and
  make it time-varying (non-stationary), instead of the stationary Matern
  space-time structure that `Z1` alone carries.
- The product term `Z_T2 * Z1` has variance `Var(Z_T2)Var(Z1)` only under
  independence, and its marginal second moment does not match a unit-variance
  Gaussian. The cross-variable and temporal properties of the sum/product no
  longer correspond to the target covariance parameters.
- No re-centering or unit-scaling step is applied, so the resulting `Z` does not
  meet the zero-mean / unit-variance latent-field requirement of
  `MSTWeatherGen_Estim`.

## Conclusion / recommendation

`Z1` alone (from `SimulateParsimNS`) is the natural candidate for a correct
latent Gaussian field: it has the target multivariate space-time structure and
is already (approximately) standardized. The extra `Z_T2`, `Z_T1` terms were
likely added to inject a temporal/non-separable component, but as written they:

1. de-standardize the margins (non-zero mean, sd ≠ 1),
2. shrink all temporal correlations below the `exp(-r_t)` target,
3. distort the cross-variable correlations away from 0.7,
4. make the field non-stationary in space due to site-common factors.

Recommended fixes, in order of preference:

- **Use `Z1` directly** as the latent field (drop `Z_T2`, `Z_T1`), or
- post-standardize `Z` (center/scale each margin to mean 0, sd 1) after
  construction, and explicitly check that the empirical temporal and
  cross-variable correlations match the targets, and
- if a pure-temporal/shared component is really wanted, build it *between*
  variables rather than as a site-common factor, and scale the combination
  analytically so the marginal variance is 1.

## Session info

```{r session, eval=TRUE}
sessionInfo()
```
