---
title: "Estimating MSTWeatherGen from a Z1 simulation"
description: "Load, inspect, plot, and estimate a non-seasonal MSTWeatherGen model from Z1"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimating MSTWeatherGen from a Z1 simulation}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

> Le plan : Introduction, generation de Z1, lancer l'estimation du modèle, comparer les paramètres et valeurs retourné. 

## Introduction

The goal of this vignette is to test the estimation phase of MSTWeatherGen and especially the function estimation_gf 
in GRF_helper_function which is an internal function 
Here is how to load it : MSTWeatherGen:::estimation_gf() when you are going to call it after in the vignette.
So you need to load the package first.

```{r load-package, message=FALSE, warning=FALSE}
library(MSTWeatherGen)
source(file.path("R", "NSMuST.R"))
devtools::load_all()
setCores(1)
```


> warning utilisation de fonction interne au package + temps d'execution

This vignette shows how to load a simulated three-dimensional array named
`Z1`, inspect its structure, plot one of its variables through time, and fit a
non-seasonal `MSTWeatherGen` model. It uses only the simulated `Z1` dataset and
the 108 spatial coordinates supplied with the package. 

The expected organization of `Z1` is:

`Z1[time, spatial location, variable]`

Here the dimensions are like the original vignette 1,461 days, 108
spatial locations, and 3 variables.

## How to generate Z1

## Generate Z1
Not to be touched by the user, necessary functions and data.
```{r , eval=FALSE}
# --- Coordonnées spatiales : 108 sites du package ---
data("coordinates", package = "MSTWeatherGen")
spatial_coordinates <- as.matrix(coordinates[, c("longitude", "latitude")])
ns <- nrow(spatial_coordinates)

# --- Fonctions de covariance Matérn (une par variable) ---
matern_fun1 <- function(t_val, s_coord) list(nu = nu[1], r = rs[1])
matern_fun2 <- function(t_val, s_coord) list(nu = nu[2], r = rs[2])
matern_fun3 <- function(t_val, s_coord) list(nu = nu[3], r = rs[3])
SpectralPars <- list(matern_fun1, matern_fun2, matern_fun3)

# --- Corrélation croisée entre variables ---
Sigma_fun <- function(t_val, s_coord) {
  M <- matrix(0.85, ncol = p, nrow = p)
  diag(M) <- 1
  M
}

# --- Anisotropie (cas isotrope) ---
aniso_fun1 <- aniso_fun2 <- aniso_fun3 <- function(t_val, s_coord) diag(2)
anisotropies <- list(aniso_fun1, aniso_fun2, aniso_fun3)


```

This chunk is the one to be modified by the user to generate the Z1 Gaussian data with known parameters that will be put in the entry of the estimation_gf and then we are going to compare the parameters obtained with the estimation and see the difference
```{r , eval=FALSE}
# --- Nombre de variables a estimer ---
p <- 3

# --- Séquence temporelle : 2018-01-01 à 2021-12-31 (1461 jours) ici changer si necessaie ---
temporal_resolution <- 1
t_seq <- seq(1, 1461, by = temporal_resolution)
dates_Z1 <- seq(as.Date("2018-01-01"), as.Date("2021-12-31"), by = "day")

# For the whole data range
#t_seq <- seq(1, 1461, by = temporal_resolution)
#dates_Z1 <- seq(as.Date("2011-01-01"), as.Date("2021-12-31"), by = "day")
nt <- length(t_seq)

# --- Parameters to change by the user ---
L  <- 3000            # nombre d'ondes (méthode spectrale)
rs <- c(0.5, 1, 2)     # range Matérn (a_i)
nu <- c(0.5, 1, 2)     # smooth Matérn (nu_i)

rho_1 <- matrix(0.85, nrow = p, ncol = p)
diag(rho_1) <- 1

params = list(
  p = p,
  nt = nt,
  t = t_seq,
  A = c(0.2,0.5,0.8), # paramètres A_i
  a = 1/2,   # a=d
  alpha = 1, # b=e, with c=f=1
  delta = 0, # does not exist in MSTWeatherGen
  b = 1, # non-separability - does not exist in MSTWeatherGen
  anisotropies = anisotropies,
  SpectralPars = SpectralPars,
  Sigma_fun    = Sigma_fun,
  rs = rs,
  nu = nu,
  rho_1 = rho_1
)

```



```{r , eval=FALSE}
source(file.path("R", "NSMuST.R"))
#library(invgamma) Des fois necessaire pour generer Z1

# Set seed if you want
set.seed(14347)
# Do not load if you do
Z1 <- SimulateParsimNS(
  L = L, ns = ns, nt = nt, p = p,
  params = params,
  spatial_coordinates = spatial_coordinates,
  SpectralDensity = MSTWeatherGen:::MaternSpectralDensity3D,
  parallelize = FALSE, batch_size = 100, n_cores = 4
)
#Takes 5 five minutes approx 
```

## Run estimation
This is how to run estimation 

### Part 1 :  Read the already saved Z1 simulation data otherwise change the path

```{r eval=FALSE}
data("Z1", package = "MSTWeatherGen")

if (length(dim(Z1)) != 3L) {
  stop("Z1 must be a three-dimensional array: time, location, variable.")
}
```

### Part 2 : Define dates, coordinates, and variable names to suit the data set

```{r eval=FALSE}
data("coordinates", package = "MSTWeatherGen")
dates <- seq(as.Date("2018-01-01"), as.Date("2021-12-31"), by = "day")
names <- c("Precipitation", "Wind", "Temp_max")
coordinates_Z1 <- coordinates[, c("longitude", "latitude")]

dimnames(Z1)[[3]] <- names

```

## Estimate and mean of many simulations

After running one of the estimation chunks, inspect and save the resulting
object as follows:

## PArt 1 Preambule function and parameters necesary for the estimation. Chunk 1

```{r eval=FALSE}
seasons <- list(
  s1 = list(min_day = 1, max_day = 29, min_month = 12, max_month = 2),
  s2 = list(min_day = 1, max_day = 31, min_month = 3, max_month = 5),
  s3 = list(min_day = 1, max_day = 31, min_month = 6, max_month = 8),
  s4 = list(min_day = 1, max_day = 30, min_month = 9, max_month = 11)
)

 # --- Paramètres ---
n1     <- 3
n2     <- 3
tmax   <- 1
max_it <- 100
season <- seasons[["s1"]]

# --- Filtrer les données pour la saison ---
filtered <- filter_season_data(data, dates, season, names)
data_s   <- filtered$data_filtered
dates_s  <- filtered$dates_filtered
rm(filtered)

# --- Types de temps ---
wt_obj <- weather_types(
  data          = data_s,
  variables     = names,
  dates         = dates_s,
  coordinates   = coordinates,
  max_number_wt = 6,
  return_plots  = FALSE
)
wt <- wt_obj$cluster
K  <- length(unique(wt))

lmbd             <- estimate_lambda_transformations(
  data = data_s, wt = wt, names = names, coordinates = coordinates)
threshold_precip <- lmbd$threshold_precip
lmbd             <- lmbd$lambda_transformations
data_s           <- transformations(
  data = data_s, wt = wt, names = names, coordinates = coordinates, lmbd = lmbd)

dst <- sapply(1:nrow(coordinates), function(i) {
  sapply(1:nrow(coordinates), function(j) {
    geosphere::distHaversine(coordinates[i,], coordinates[j,]) / 1000
  })
})

cr <- sapply(names, function(v1) {
  sapply(names, function(v2) {
    mean(sapply(1:dim(data_s)[2], function(j)
      cor(data_s[,j,v1], data_s[,j,v2], use="complete.obs")), na.rm=TRUE)
  })
})
cr <- matrix(cr, nrow=length(names), ncol=length(names))
colnames(cr) <- rownames(cr) <- names

ep <- generate_variable_index_pairs(names)
rho1_init <- data.frame(
  v1  = ep[,1],
  v2  = ep[,2],
  cov = ifelse(ep[,1] == ep[,2], 1, 0.1),
  stringsAsFactors = FALSE
)

```

## PArt 2 Lancer les simulations. 
Ici weather type environ vaut 5, donc number_simulations <- 3 implique total_sim = 3*5= 15 simulation au total. Des fois K vaut 6 ou 4.

```{r eval=FALSE}
number_simulations <- 3

# Stocker tous les résultats dans une liste nommée
# Structure : estim_results[[k]][[i]] = résultat du run i pour le type de temps k
estim_results <- vector(mode="list", length=K)

for (k in 1:K) {

  wt_id <- which(wt == k)
  wt_id <- wt_id[wt_id > tmax + 1]

  estim_results[[k]] <- vector(mode="list", length=number_simulations)
  names(estim_results[[k]]) <- paste0("run_", 1:number_simulations)

  for (i in seq_len(number_simulations)) {
    message(sprintf("  k=%d — Run %d/%d", k, i, number_simulations))
    set.seed(14347 + i)

    estim_results[[k]][[i]] <- estimation_gf(
      data             = data_s,
      wt_id            = wt_id,
      max_it           = max_it,
      dates            = dates_s,
      tmax             = tmax,
      names            = names,
      coordinates      = coordinates,
      n1               = n1,
      n2               = n2,
      rho1             = rho1_init,
      cr               = cr,
      threshold_precip = threshold_precip[[k]]
    )
  }
}

```

## Comparing parameters and estimations values

> comment on compare les résultats

```{r eval=FALSE}
# ============================================================
# PARTIE 3 — Moyennes des paramètres estimés
# ============================================================

cols_num <- c("a","b","c","d","e","Ai","Aj","aii","ajj",
              "nuii","nujj","r1ii","r1jj","r2ii","r2jj",
              "rho1ij","rho2ij")

# Moyenne sur les runs pour chaque type de temps
mean_by_wt <- lapply(1:K, function(k) {
  if (is.null(estim_results[[k]])) return(NULL)
  parm_all_runs <- do.call(rbind, lapply(estim_results[[k]], function(run) run$parm))
  aggregate(parm_all_runs[, cols_num], 
            by = list(v1=parm_all_runs$v1, v2=parm_all_runs$v2), 
            FUN = mean, na.rm = TRUE)
})
names(mean_by_wt) <- paste0("wt_", 1:K)

# Moyenne globale sur tous les types de temps et tous les runs
parm_all <- do.call(rbind, lapply(1:K, function(k) {
  if (is.null(estim_results[[k]])) return(NULL)
  do.call(rbind, lapply(estim_results[[k]], function(run) run$parm))
}))

mean_global <- aggregate(parm_all[, cols_num],
                         by = list(v1=parm_all$v1, v2=parm_all$v2),
                         FUN = mean, na.rm = TRUE)
```
