---
title: "Estimating MSTWeatherGen from a Z1 simulation"
description: "Load, inspect, plot, and estimate a non-seasonal MSTWeatherGen model from Z1"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimating MSTWeatherGen from a Z1 simulation}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

> Le plan : Introduction, generation de Z1, lancer l'estimation du modèle, comparer les paramètres et valeurs retourné.

## Introduction

The goal of this vignette is to test the estimation phase of MSTWeatherGen and especially the function estimation_gf 
in GRF_helper_function which is an internal function.  
Here is how to load the package internal function that is use.

```{r load-package, message=FALSE, warning=FALSE}
devtools::load_all()
# Set the number of cores to use
cores_nb <- setCores(1)
```

## Estimate from know parameters

Here we shows how to create a simulated three-dimensional array named
`Z1` that fit a non-seasonal `MSTWeatherGen` model estimation.  
The expected organization of `Z1` is:  
`Z1[time, spatial location, variable]`

> Here the dimensions are like the original vignette 1,461 days, 108
spatial locations, and 3 variables c("Precipitation", "Wind", "Temp_max").

### Generate Z1

This part doesn't have to be executed. `Z1` is already available as a data by the package.  
```{r , eval=FALSE}
```

This chunk is the one to be modified by the user to generate the Z1 Gaussian data with known parameters that will be put in the entry of the estimation_gf and then we are going to compare the parameters obtained with the estimation and see the difference
```{r , eval=FALSE}
# --- Nombre de variables a estimer ---

```



```{r , eval=FALSE}
#library(invgamma) may be necessary

# Set seed if you want
set.seed(14347)
# Do not load if you do

# Define spatial and temporal resolutions
spatial_resolution  = 0.4
temporal_resolution = 1

# Generate spatial grid for simulation
s1_seq = seq(0, 10, by = spatial_resolution)
s2_seq = seq(0, 8, by = spatial_resolution)
grid_spatial = expand.grid(s1 = s1_seq, s2 = s2_seq)
spatial_coordinates = as.matrix(grid_spatial) 
ns = nrow(spatial_coordinates)   

# Define temporal sequence
t_seq = seq(1, 30, by = temporal_resolution)
nt = length(t_seq)  

p = 3 # Number of variables
L = 3000 # Number of waves

# Définition des Matern
rs = c(0.5,1,2) # paramètres a dans le papier
nu = c(0.5,1,2) 

rho_1 <- matrix(0.7, nrow = p, ncol = p)
diag(rho_1) <- 1
# Define Matern covariance function parameters
matern_fun1 = function(t_val, s_coord) {
  list(nu=nu[1], r=rs[1])
}
matern_fun2 = function(t_val, s_coord) {
  list(nu=nu[2], r=rs[2])
}
matern_fun3 = function(t_val, s_coord) {
  list(nu=nu[3], r=rs[3])
}
SpectralPars = list(matern_fun1, matern_fun2, matern_fun3)


# Define covariance structure (between variables) function
Sigma_fun = function(t_val, s_coord) {
  M = matrix(0.7,ncol=3,nrow=3)
  diag(M) = 1
  return(M)
}

# Define anisotropy functions for Isotropic case
aniso_fun1 = aniso_fun2 = aniso_fun3 = function(t_val, s_coord) {
  diag(2)
}
anisotropies = list(aniso_fun1, aniso_fun2, aniso_fun3)


# Define simulation parameters
params = list(
  p = p,
  nt = nt,
  t = t_seq,
  A = c(0.2,0.5,0.8), # paramètres A_i
  a = 1/2,   # a=d
  alpha = 1, # b=e, with c=f=1
  delta = 0, # does not exist in MSTWeatherGen
  b = 1, # non-separability - does not exist in MSTWeatherGen
  anisotropies = anisotropies,
  SpectralPars = SpectralPars,
  Sigma_fun    = Sigma_fun,
  rs = rs,
  nu = nu)

# Run spatio-temporal simulation 
set.seed(14347)
Z1 = SimulateParsimNS(
  L                  = L,
  ns                 = ns,
  nt                 = nt,
  p                  = p,
  params             = params,
  spatial_coordinates = spatial_coordinates,
  SpectralDensity = MaternSpectralDensity3D,
  parallelize        = F, 
  batch_size = 100,
  n_cores = 4
)

r_t = c(0.2,0.4,0.6)
rho  = Sigma_fun(1,1)
Z_T1 = TemporalSim(nt,r_t,rho,a=rep(1,p),nu=rep(1,p),A=rep(0,p))
Z_T2 = TemporalSim(nt,r_t,rho,a=rs,nu=nu,A=c(0.2,0.5,0.8))

Z = Z1
for (i in seq_len(ns)) {
  for (k in seq_len(p)){
    Z[,i,k] = 0.707*(Z_T2[,k]*Z1[,i,k] + Z_T1[,k])
  }
}

#Takes 5 five minutes to run
```

### Run estimation  
This is how to run estimation 

#### Part 1 : Read the already saved Z1 simulation data 

```{r eval=FALSE}
data("Z", package = "MSTWeatherGen")

if (length(dim(Z)) != 3L) {
  stop("Z must be a three-dimensional array: time, location, variable.")
}
```

#### Part 2 : Define dates, coordinates, and variable names to suit the data set

```{r eval=FALSE}
wt_id = 5:(dim(Z)[1]-5)
tmax   <- 3   
max_it <- 100 # Maximum number of iteration for optimisation process

dates = seq(as.Date('2018-01-01'),as.Date('2018-01-30'),by = 'day')
names = c('v1','v2','v3')
n1 = 10
n2 = 10
spatial_coordinates = as.data.frame(spatial_coordinates)
coordinates = spatial_coordinates
names(coordinates) = c('longitude','latitude')

rho1_init <- matrix(0.1, nrow = p, ncol = p)
diag(rho1_init) <- 0.5
dimnames(Z)[[3]] = names

cr <- sapply(names, function(v1) {
  sapply(names, function(v2) {
    mean(sapply(1:dim(Z)[2], function(j)
      cor(Z[,j,v1], Z[,j,v2], use="complete.obs")), na.rm=TRUE)
  })
})
cr <- matrix(cr, nrow=length(names), ncol=length(names))
colnames(cr) <- rownames(cr) <- names
```

#### Part 3 Preambule function and parameters necesary for the estimation.

```{r eval=FALSE}
```

#### Part 4 Simulations. 

Ici weather type environ vaut 6, donc number_simulations <- 3 implique total_sim = 3*6= 18 simulation au total. Des fois K vaut 4 ou 5.

```{r eval=FALSE}
estim_results <- estimation_gf(
      data             = Z,
      wt_id            = wt_id,
      max_it           = max_it,
      dates            = dates,
      tmax             = tmax,
      names            = names,
      coordinates      = coordinates,
      n1               = n1,
      n2               = n2,
      rho1             = rho1_init,
      cr               = cr,
      threshold_precip = threshold_precip[[k]])


# Moyenner les simulations

cols_num <- c("a","b","c","d","e","Ai","Aj","aii","ajj",
              "nuii","nujj","r1ii","r1jj","r2ii","r2jj",
              "rho1ij","rho2ij")

# Moyenne sur les runs pour chaque type de temps
mean_by_wt <- lapply(1:K, function(k) {
  if (is.null(estim_results[[k]])) return(NULL)
  parm_all_runs <- do.call(rbind, lapply(estim_results[[k]], function(run) run))
  aggregate(parm_all_runs[, cols_num], 
            by = list(v1=parm_all_runs$v1, v2=parm_all_runs$v2), 
            FUN = mean, na.rm = TRUE)
})
names(mean_by_wt) <- paste0("wt_", 1:K)

# Moyenne globale sur tous les types de temps et tous les runs
parm_all <- do.call(rbind, lapply(mean_by_wt, function(w) w))
mean_global <- aggregate(parm_all[, cols_num],
                         by = list(v1=parm_all$v1, v2=parm_all$v2),
                         FUN = mean, na.rm = TRUE)
```


## Comparing parameters and estimations values

On met les parametres initiaux dans un dataframe qui ressemble a mean global
```{r eval=FALSE}
# --- Table des vraies valeurs, structurée comme mean_global ---

# Toutes les paires (v1, v2), diagonale + hors-diagonale, comme dans mean_global
var_pairs <- expand.grid(v1 = names, v2 = names, stringsAsFactors = FALSE)
var_pairs <- var_pairs[as.character(var_pairs$v1) <= as.character(var_pairs$v2), ]

# Indices correspondants dans rs / nu / A (dans l'ordre de `names`)
i <- match(var_pairs$v1, names)
j <- match(var_pairs$v2, names)

true_params <- data.frame(
  v1     = var_pairs$v1,
  v2     = var_pairs$v2,
  a      = params$a,      
  b      = params$alpha,   
  c      = 1,              
  d      = params$a,       
  e      = params$alpha,   
  Ai     = params$A[i],
  Aj     = params$A[j],
  aii    = params$rs[i],
  ajj    = params$rs[j],
  nuii   = params$nu[i],
  nujj   = params$nu[j],
  rho1ij = mapply(function(a, b) rho_1[a, b], i, j)
)

true_params #les params mis dans Z1 mais sous forme df
#avoir les df cote a cote
mean_global
```

## Plotting

Plotting with know value (original Z1 parameters)

```{r eval=FALSE}
nb_points = 30 #Si plus risque d'etre illisible

nt <- length(dates)
r_t <- c(0.2,0.4,0.6)
rho_1 <- Sigma_fun(1,1)
rs <- c(0.5, 1, 2)     # range Matérn (a_ii)
nu <- c(0.5, 1, 2)     # smooth Matérn (nu_ii)
Z_T1 <- TemporalSim(nt,r_t,rho,a=rs,nu=nu,A=c(0.2,0.5,0.8))
plot(1:nb_points, Z_T1[1:nb_points, 1], type = "l")
points(1:nb_points, Z_T1[1:nb_points, 2], type = "l", lty = 2)
points(1:nb_points, Z_T1[1:nb_points, 3], type = "l", lty = 3)
```

Same plot

```{r eval=FALSE}
#Same things different way

# --- Récupérer les vrais paramètres par variable (diagonale v1 == v2) ---
diag_true <- true_params[true_params$v1 == true_params$v2, ]
diag_true <- diag_true[match(names, diag_true$v1), ]  

rs_true <- diag_true$aii
nu_true <- diag_true$nuii
A_true  <- diag_true$Ai
rho_1  = Sigma_fun(1,1)

# --- Trajectoire temporelle avec les vrais paramètres ---
Z_T2_true <- TemporalSim(nt, r_t, rho_1, a = rs_true, nu = nu_true, A = A_true)

plot(1:nb_points, Z_T2_true[1:nb_points, 1], type = "l")
points(1:nb_points, Z_T2_true[1:nb_points, 2], type = "l", lty = 2)
points(1:nb_points, Z_T2_true[1:nb_points, 3], type = "l", lty = 3)
```

Plot with estimated parameters
```{r eval=FALSE}
# --- Récupérer les paramètres estimés par variable (diagonale v1 == v2) ---
diag_est <- mean_global[mean_global$v1 == mean_global$v2, ]
diag_est <- diag_est[match(names, diag_est$v1), ]  # remettre dans l'ordre de `names`

rs_est <- diag_est$aii
nu_est <- diag_est$nuii
A_est  <- diag_est$Ai

rho1_est <- sapply(names, function(v1) {
          sapply(names, function(v2) {
      rho1ij <- mean_global[mean_global$v1 == v1 & mean_global$v2 == v2 | mean_global$v1 == v2 & mean_global$v2 == v1,]$rho1ij
      return(rho1ij)
    })
  })
# --- Trajectoire temporelle avec les paramètres estimés ---
Z_T2_est <- TemporalSim(nt, r_t, rho1_est, a = rs_est, nu = nu_est, A = A_est)

plot(1:nb_points, Z_T2_est[1:nb_points, 1], type = "l")
points(1:nb_points, Z_T2_est[1:nb_points, 2], type = "l", lty = 2)
points(1:nb_points, Z_T2_est[1:nb_points, 3], type = "l", lty = 3)
```

## Plotting at time 1 precipitation and Temp_max

```{r eval=FALSE}
ns <- nrow(spatial_coordinates)

Z = Z1
for (i in seq_len(ns)){
  for (k in seq_len(p)){
    Z[,i,k] = Z_T2_est[,k]*Z1[,i,k] + Z_T1[,k]
  }
}

# Plot the variable
x_coords = spatial_coordinates[, 1]
y_coords = spatial_coordinates[, 2]

df_v1 = data.frame(
  ZZ1      = Z[1, , 2],
  X        = x_coords,
  Y        = y_coords,
  Variable = paste("V1"),
  Time     =  1
)

df_v2 = data.frame(
  ZZ2      = Z[1, , 3],
  X        = x_coords,
  Y        = y_coords,
  Variable = paste("V3"),
  Time     =  1
)


df_list = list()
df_list[[length(df_list) + 1]] = df_v1
df_list[[length(df_list) + 1]] = df_v2
df = dplyr::bind_rows(df_list)
df = df %>% dplyr::mutate(Time = "Time 1")

time_steps_vec = 1
df_var1 = df %>% filter(Variable == "V1")
df_var2 = df %>% filter(Variable == "V3")

color_scale_fun = function(...) scale_fill_viridis_c(option = "H", ...)
theme_fun = theme_bw

p_1 = ggplot(df_var1, aes(x = X, y = Y, fill = ZZ1)) +
  geom_tile(width = 0.1, height = 0.1) +
  color_scale_fun(name = "Precipitation") +
  #facet_wrap(~ Time, nrow = 1) +
  theme_fun() +
  labs(
    title = paste("Precipitation ; time=1"),
    x = "x coordinate",
    y = "y coordinate",
    fill = NULL
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12),
    strip.text = element_text(size = 10)
  )


p_2 = ggplot(df_var2, aes(x = X, y = Y, fill = ZZ2)) +
  #geom_raster() +
  geom_tile(width = 0.1, height = 0.1) +
  color_scale_fun(name = "Temp_max") +
  #facet_wrap(~ Time, nrow = 1) +
  theme_fun() +
  labs(
    title = paste("Temp_max ; time = 1"),
    x = "x coordinate",
    y = "y coordinate",
    fill = NULL
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12),
    strip.text = element_text(size = 10)
  ) 
pl = p_1 + p_2 + plot_layout(guides = 'collect')
pl
```
