---
title: "Estimating MSTWeatherGen from a Z1 simulation"
description: "Load, inspect, plot, and estimate a non-seasonal MSTWeatherGen model from Z1"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimating MSTWeatherGen from a Z1 simulation}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This vignette shows how to load a simulated three-dimensional array named
`Z1`, inspect its structure, plot one of its variables through time, and fit a
non-seasonal `MSTWeatherGen` model. It uses only the simulated `Z1` dataset and
the 108 spatial coordinates supplied with the package. 

The expected organization of `Z1` is:

`Z1[time, spatial location, variable]`

Here the dimensions are like the original vignette 1,461 days, 108
spatial locations, and 3 variables.

## Load MSTWeatherGen

```{r load-package, message=FALSE, warning=FALSE}
library(MSTWeatherGen)
```

## Part 1 :  Read the Z1 simulation data

The code below searches for `Z1_V2.rds` from either the package source tree or
the local project. For an installed package, the recommended location is
`inst/extdata/Z1_V2.rds`.

```{r eval=FALSE}
fichier_Z1 <- "/home/aboualam/MSTWeatherGen/data/Z1_v3.rds"
Z1 <- readRDS(fichier_Z1)

if (length(dim(Z1)) != 3L) {
  stop("Z1 must be a three-dimensional array: time, location, variable.")
}
# Meme dimension que data utilise dans la vignette originale.
```

## Part 2 : Define dates, coordinates, and variable names

```{r eval=FALSE}
data("coordinates", package = "MSTWeatherGen")
dates <- seq(as.Date("2018-01-01"), as.Date("2021-12-31"), by = "day")
names <- c("Precipitation", "Wind", "Temp_max")
coordinates_Z1 <- coordinates[, c("longitude", "latitude")]

dimnames(Z1)[[3]] <- names

```


## Plot one variable through time A reverifier et pas obligatoire dans la vignette. Signal gaussien observe.

The following function plots one variable at one spatial location. The
defaults display the third variable (`Temp_max`) at the first location.

```{r eval=FALSE}
plot_Z1_variable <- function(
  Z,
  dates,
  variable = 3,
  spatial_position = 1,
  variable_names = dimnames(Z)[[3]]
) {
  variable_label <- if (is.null(variable_names)) {
    paste("Variable", variable)
  } else {
    variable_names[variable]
  }

  plot(
    dates,
    Z[, spatial_position, variable],
    type = "l",
    lwd = 1,
    col = "steelblue4",
    xlab = "Date",
    ylab = variable_label,
    main = paste(variable_label, "- spatial location", spatial_position)
  )
}
variable = 1
# Plotter une variable i entre 1 et 3 par rapport au temps ici precipitation vs
plot_Z1_variable(Z = Z1, dates = dates, variable = variable, spatial_position = 1)
```

## Estimate a non-seasonal model

The estimation is performed over the full 2018--2021 period. Consequently,
`by_season` is set to `FALSE` and no `seasons` argument is supplied.

`scale` is also set to `FALSE`: the internal seasonal standardization requires
more than five years of observations, whereas this example contains four
years.

The following computation can take time, so this chunk is not run while the
vignette is built. Remove `eval=FALSE` or run the code interactively to perform
the estimation.

```{r eval=FALSE}
seasons <- list(
  s1 = list(min_day = 1, max_day = 29, min_month = 12, max_month = 2),
  s2 = list(min_day = 1, max_day = 31, min_month = 3, max_month = 5),
  s3 = list(min_day = 1, max_day = 31, min_month = 6, max_month = 8),
  s4 = list(min_day = 1, max_day = 30, min_month = 9, max_month = 11)
)

swg_Z1 <- MSTWeatherGen_Estim(
  data                = Z1,
  seasons             = seasons,
  dates               = dates,
  names               = names,
  by_season           = TRUE,
  scale               = FALSE,
  precipitation       = FALSE,
  names_weather_types = names,
  coordinates         = coordinates,
  max_it              = 100,
  tmax                 = 1,
  n1                   = 3,
  n2                   = 3
)
```

## Estimate and mean of many simulations

After running one of the estimation chunks, inspect and save the resulting
object as follows:

## Preambule


```{r eval=FALSE}
 # --- Paramètres ---
n1     <- 3
n2     <- 3
tmax   <- 1
max_it <- 100
season <- seasons[["s1"]]

# --- Filtrer les données pour la saison ---
filtered <- filter_season_data(data, dates, season, names)
data_s   <- filtered$data_filtered
dates_s  <- filtered$dates_filtered
rm(filtered)

# --- Types de temps ---
wt_obj <- weather_types(
  data          = data_s,
  variables     = names,
  dates         = dates_s,
  coordinates   = coordinates,
  max_number_wt = 6,
  return_plots  = FALSE
)
wt <- wt_obj$cluster
K  <- length(unique(wt))

lmbd             <- estimate_lambda_transformations(
  data = data_s, wt = wt, names = names, coordinates = coordinates)
threshold_precip <- lmbd$threshold_precip
lmbd             <- lmbd$lambda_transformations
data_s           <- transformations(
  data = data_s, wt = wt, names = names, coordinates = coordinates, lmbd = lmbd)

dst <- sapply(1:nrow(coordinates), function(i) {
  sapply(1:nrow(coordinates), function(j) {
    geosphere::distHaversine(coordinates[i,], coordinates[j,]) / 1000
  })
})

cr <- sapply(names, function(v1) {
  sapply(names, function(v2) {
    mean(sapply(1:dim(data_s)[2], function(j)
      cor(data_s[,j,v1], data_s[,j,v2], use="complete.obs")), na.rm=TRUE)
  })
})
cr <- matrix(cr, nrow=length(names), ncol=length(names))
colnames(cr) <- rownames(cr) <- names

ep <- generate_variable_index_pairs(names)
rho1_init <- data.frame(
  v1  = ep[,1],
  v2  = ep[,2],
  cov = ifelse(ep[,1] == ep[,2], 1, 0.1),
  stringsAsFactors = FALSE
)

```

## Lancer les simulations

```{r eval=FALSE}
number_simulations <- 3

# Stocker tous les résultats dans une liste nommée
# Structure : estim_results[[k]][[i]] = résultat du run i pour le type de temps k
estim_results <- vector(mode="list", length=K)

for (k in 1:K) {

  wt_id <- which(wt == k)
  wt_id <- wt_id[wt_id > tmax + 1]

  estim_results[[k]] <- vector(mode="list", length=number_simulations)
  names(estim_results[[k]]) <- paste0("run_", 1:number_simulations)

  for (i in seq_len(number_simulations)) {
    message(sprintf("  k=%d — Run %d/%d", k, i, number_simulations))
    set.seed(14347 + i)

    estim_results[[k]][[i]] <- estimation_gf(
      data             = data_s,
      wt_id            = wt_id,
      max_it           = max_it,
      dates            = dates_s,
      tmax             = tmax,
      names            = names,
      coordinates      = coordinates,
      n1               = n1,
      n2               = n2,
      rho1             = rho1_init,
      cr               = cr,
      threshold_precip = threshold_precip[[k]]
    )
}

```


```{r eval=FALSE}
# ============================================================
# PARTIE 3 — Moyennes des paramètres estimés
# ============================================================

cols_num <- c("a","b","c","d","e","Ai","Aj","aii","ajj",
              "nuii","nujj","r1ii","r1jj","r2ii","r2jj",
              "rho1ij","rho2ij")

# Moyenne sur les runs pour chaque type de temps
mean_by_wt <- lapply(1:K, function(k) {
  if (is.null(estim_results[[k]])) return(NULL)
  parm_all_runs <- do.call(rbind, lapply(estim_results[[k]], function(run) run$parm))
  aggregate(parm_all_runs[, cols_num], 
            by = list(v1=parm_all_runs$v1, v2=parm_all_runs$v2), 
            FUN = mean, na.rm = TRUE)
})
names(mean_by_wt) <- paste0("wt_", 1:K)

# Moyenne globale sur tous les types de temps et tous les runs
parm_all <- do.call(rbind, lapply(1:K, function(k) {
  if (is.null(estim_results[[k]])) return(NULL)
  do.call(rbind, lapply(estim_results[[k]], function(run) run$parm))
}))

mean_global <- aggregate(parm_all[, cols_num],
                         by = list(v1=parm_all$v1, v2=parm_all$v2),
                         FUN = mean, na.rm = TRUE)
```
