---
title: "Estimating MSTWeatherGen from a Z1 simulation"
description: "Load, inspect, plot, and estimate a non-seasonal MSTWeatherGen model from Z1"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimating MSTWeatherGen from a Z1 simulation}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

> Le plan : Introduction, generation de Z1, lancer l'estimation du modèle, comparer les paramètres et valeurs retourné. 

## Introduction

The goal of this vignette is to test the estimation phase of MSTWeatherGen and especially the function estimation_gf 
in GRF_helper_function which is an internal function.  
Here is how to load the package internal function that is use.

```{r load-package, message=FALSE, warning=FALSE}
devtools::load_all()
# Set the number of cores to use
cores_nb <- setCores(1)
```

## Estimate from know parameters

Here we shows how to create a simulated three-dimensional array named
`Z1` that fit a non-seasonal `MSTWeatherGen` model estimation.  
The expected organization of `Z1` is:  
`Z1[time, spatial location, variable]`

> Here the dimensions are like the original vignette 1,461 days, 108
spatial locations, and 3 variables c("Precipitation", "Wind", "Temp_max").

### Generate Z1

This part doesn't have to be executed. `Z1` is already available as a data by the package.  
```{r , eval=FALSE}
# --- Coordonnées spatiales : 108 sites du package ---
data("coordinates", package = "MSTWeatherGen")
spatial_coordinates <- as.matrix(coordinates[, c("longitude", "latitude")])
ns <- nrow(spatial_coordinates)

# --- Fonctions de covariance Matérn (une par variable) ---
matern_fun1 <- function(t_val, s_coord) list(nu = nu[1], r = rs[1])
matern_fun2 <- function(t_val, s_coord) list(nu = nu[2], r = rs[2])
matern_fun3 <- function(t_val, s_coord) list(nu = nu[3], r = rs[3])
SpectralPars <- list(matern_fun1, matern_fun2, matern_fun3)

# --- Corrélation croisée entre variables ---
Sigma_fun <- function(t_val, s_coord) {
  M <- matrix(0.85, ncol = p, nrow = p)
  diag(M) <- 1
  M
}

# --- Anisotropie (cas isotrope) ---
aniso_fun1 <- aniso_fun2 <- aniso_fun3 <- function(t_val, s_coord) diag(2)
anisotropies <- list(aniso_fun1, aniso_fun2, aniso_fun3)


```

This chunk is the one to be modified by the user to generate the Z1 Gaussian data with known parameters that will be put in the entry of the estimation_gf and then we are going to compare the parameters obtained with the estimation and see the difference
```{r , eval=FALSE}
# --- Nombre de variables a estimer ---
p <- 3

# --- Séquence temporelle : 2018-01-01 à 2021-12-31 (1461 jours) ici changer si necessaie ---
# the whole dates availables
#dates_Z1 <- seq(as.Date("2011-01-01"), as.Date("2021-12-31"), by = "day")
# Part of the whole dates data
dates_Z1 <- seq(as.Date("2018-01-01"), as.Date("2021-12-31"), by = "day")
nt <- length(dates_Z1)
temporal_resolution <- 1
t_seq <- seq(1, nt, by = temporal_resolution)

# --- Parameters to change by the user ---
L  <- 3000             # nombre d'ondes (méthode spectrale)
rs <- c(0.5, 1, 2)     # range Matérn (a_ii)
nu <- c(0.5, 1, 2)     # smooth Matérn (nu_ii)

rho_1 <- matrix(0.85, nrow = p, ncol = p)
diag(rho_1) <- 1

params = list(
  p = p,
  nt = nt,
  t = t_seq,
  A = c(0.2,0.5,0.8), # paramètres A_i
  a = 1/2,   # a=d
  alpha = 1, # b=e, with c=f=1
  delta = 0, # does not exist in MSTWeatherGen
  b = 1, # non-separability - does not exist in MSTWeatherGen
  anisotropies = anisotropies,
  SpectralPars = SpectralPars,
  Sigma_fun    = Sigma_fun,
  rs = rs,
  nu = nu,
  rho_1 = rho_1
)

```



```{r , eval=FALSE}
#library(invgamma) may be necessary

# Set seed if you want
set.seed(14347)
# Do not load if you do
Z1 <- SimulateParsimNS(
  L = L, ns = ns, nt = nt, p = p,
  params = params,
  spatial_coordinates = spatial_coordinates,
  SpectralDensity = MSTWeatherGen:::MaternSpectralDensity3D,
  parallelize = FALSE, batch_size = 100, n_cores = cores_nb
)
#Takes 5 five minutes to run
```

### Run estimation  
This is how to run estimation 

#### Part 1 : Read the already saved Z1 simulation data 

```{r eval=FALSE}
data("Z1", package = "MSTWeatherGen")

if (length(dim(Z1)) != 3L) {
  stop("Z1 must be a three-dimensional array: time, location, variable.")
}
```

#### Part 2 : Define dates, coordinates, and variable names to suit the data set

```{r eval=FALSE}
data("coordinates", package = "MSTWeatherGen")
dates <- seq(as.Date("2011-01-01"), as.Date("2021-12-31"), by = "day")
dates_used <- seq(as.Date("2018-01-01"), as.Date("2021-12-31"), by = "day")
index_dates <- dates %in% dates_used
data <- data[index_dates,,]
dates <- dates_used
names <- c("Precipitation", "Wind", "Temp_max")
coordinates_Z1 <- coordinates[, c("longitude", "latitude")]

dimnames(Z1)[[3]] <- names

```

#### Part 3 Preambule function and parameters necesary for the estimation.

```{r eval=FALSE}
 # --- Paramètres ---
season <- list(min_day = 1, max_day = 31, min_month = 1, max_month = 12)  # One season - 1 year
n1     <- 3   # The number of points to uniformly distribute across the spatial domain.
n2     <- 3   # The number of points to select based on proximity to each of the n1 points.
tmax   <- 1   # Maximun time lag
max_it <- 100 # Maximum number of iteration for optimisation process

# --- Types de temps ---
wt_obj <- weather_types(
  data          = data,
  variables     = names,
  dates         = dates,
  coordinates   = coordinates,
  max_number_wt = 6,
  return_plots  = FALSE
)
wt <- wt_obj$cluster
K  <- length(unique(wt))

lmbd <- estimate_lambda_transformations(
  data = data, wt = wt, names = names,
  coordinates = coordinates)
threshold_precip <- lmbd$threshold_precip
lmbd             <- lmbd$lambda_transformations
data_t <- transformations(
  data = data, wt = wt, names = names, coordinates = coordinates, lmbd = lmbd)

cr <- sapply(names, function(v1) {
  sapply(names, function(v2) {
    mean(sapply(1:dim(data_t)[2], function(j)
      cor(data_t[,j,v1], data_t[,j,v2], use="complete.obs")), na.rm=TRUE)
  })
})
cr <- matrix(cr, nrow=length(names), ncol=length(names))
colnames(cr) <- rownames(cr) <- names

p <- 3 # nb of variables
rho1_init <- matrix(0.85, nrow = p, ncol = p)
diag(rho1_init) <- 1
```

#### Part 4 Simulations. 

Ici weather type environ vaut 6, donc number_simulations <- 3 implique total_sim = 3*6= 18 simulation au total. Des fois K vaut 4 ou 5.

```{r eval=FALSE}
number_simulations <- 2

# Stocker tous les résultats dans une liste nommée
# Structure : estim_results[[k]][[i]] = résultat du run i pour le type de temps k
estim_results <- vector(mode="list", length=K)

for (k in 1:K) {

  wt_id <- which(wt == k)
  wt_id <- wt_id[wt_id > tmax + 1]

  estim_results[[k]] <- vector(mode="list", length=number_simulations)
  names(estim_results[[k]]) <- paste0("run_", 1:number_simulations)

  for (i in seq_len(number_simulations)) {
    message(sprintf("  k=%d — Run %d/%d", k, i, number_simulations))
    set.seed(14347 + i)

    estim_results[[k]][[i]] <- estimation_gf(
      data             = data_t,
      wt_id            = wt_id,
      max_it           = max_it,
      dates            = dates,
      tmax             = tmax,
      names            = names,
      coordinates      = coordinates,
      n1               = n1,
      n2               = n2,
      rho1             = rho1_init,
      cr               = cr,
      threshold_precip = threshold_precip[[k]]
    )$parm
  }
}

save(estim_results, file="estimation_Z1.rda")

# Moyenner les simulations

cols_num <- c("a","b","c","d","e","Ai","Aj","aii","ajj",
              "nuii","nujj","r1ii","r1jj","r2ii","r2jj",
              "rho1ij","rho2ij")

# Moyenne sur les runs pour chaque type de temps
mean_by_wt <- lapply(1:K, function(k) {
  if (is.null(estim_results[[k]])) return(NULL)
  parm_all_runs <- do.call(rbind, lapply(estim_results[[k]], function(run) run))
  aggregate(parm_all_runs[, cols_num], 
            by = list(v1=parm_all_runs$v1, v2=parm_all_runs$v2), 
            FUN = mean, na.rm = TRUE)
})
names(mean_by_wt) <- paste0("wt_", 1:K)

# Moyenne globale sur tous les types de temps et tous les runs
parm_all <- do.call(rbind, lapply(mean_by_wt, function(w) w))
mean_global <- aggregate(parm_all[, cols_num],
                         by = list(v1=parm_all$v1, v2=parm_all$v2),
                         FUN = mean, na.rm = TRUE)
```


## Comparing parameters and estimations values

On met les parametres initiaux dans un dataframe qui ressemble a mean global
```{r eval=FALSE}
# --- Table des vraies valeurs, structurée comme mean_global ---

# Toutes les paires (v1, v2), diagonale + hors-diagonale, comme dans mean_global
var_pairs <- expand.grid(v1 = names, v2 = names, stringsAsFactors = FALSE)
var_pairs <- var_pairs[as.character(var_pairs$v1) <= as.character(var_pairs$v2), ]

# Indices correspondants dans rs / nu / A (dans l'ordre de `names`)
i <- match(var_pairs$v1, names)
j <- match(var_pairs$v2, names)

true_params <- data.frame(
  v1     = var_pairs$v1,
  v2     = var_pairs$v2,
  a      = params$a,      
  b      = params$alpha,   
  c      = 1,              
  d      = params$a,       
  e      = params$alpha,   
  Ai     = params$A[i],
  Aj     = params$A[j],
  aii    = params$rs[i],
  ajj    = params$rs[j],
  nuii   = params$nu[i],
  nujj   = params$nu[j],
  rho1ij = mapply(function(a, b) rho_1[a, b], i, j)
)

true_params #les params mis dans Z1 mais sous forme df
#avoir les df cote a cote
mean_global
```

## Plotting

Plotting with know value (original Z1 parameters)

```{r eval=FALSE}
nb_points = 30 #Si plus risque d'etre illisible

nt <- length(dates)
r_t <- c(0.2,0.4,0.6)
rho_1 <- Sigma_fun(1,1)
rs <- c(0.5, 1, 2)     # range Matérn (a_ii)
nu <- c(0.5, 1, 2)     # smooth Matérn (nu_ii)
Z_T1 <- TemporalSim(nt,r_t,rho,a=rs,nu=nu,A=c(0.2,0.5,0.8))
plot(1:nb_points, Z_T1[1:nb_points, 1], type = "l")
points(1:nb_points, Z_T1[1:nb_points, 2], type = "l", lty = 2)
points(1:nb_points, Z_T1[1:nb_points, 3], type = "l", lty = 3)
```

Same plot

```{r eval=FALSE}
#Same things different way

# --- Récupérer les vrais paramètres par variable (diagonale v1 == v2) ---
diag_true <- true_params[true_params$v1 == true_params$v2, ]
diag_true <- diag_true[match(names, diag_true$v1), ]  

rs_true <- diag_true$aii
nu_true <- diag_true$nuii
A_true  <- diag_true$Ai
rho_1  = Sigma_fun(1,1)

# --- Trajectoire temporelle avec les vrais paramètres ---
Z_T2_true <- TemporalSim(nt, r_t, rho_1, a = rs_true, nu = nu_true, A = A_true)

plot(1:nb_points, Z_T2_true[1:nb_points, 1], type = "l")
points(1:nb_points, Z_T2_true[1:nb_points, 2], type = "l", lty = 2)
points(1:nb_points, Z_T2_true[1:nb_points, 3], type = "l", lty = 3)
```

Plot with estimated parameters
```{r eval=FALSE}
# --- Récupérer les paramètres estimés par variable (diagonale v1 == v2) ---
diag_est <- mean_global[mean_global$v1 == mean_global$v2, ]
diag_est <- diag_est[match(names, diag_est$v1), ]  # remettre dans l'ordre de `names`

rs_est <- diag_est$aii
nu_est <- diag_est$nuii
A_est  <- diag_est$Ai

rho1_est <- sapply(names, function(v1) {
          sapply(names, function(v2) {
      rho1ij <- mean_global[mean_global$v1 == v1 & mean_global$v2 == v2 | mean_global$v1 == v2 & mean_global$v2 == v1,]$rho1ij
      return(rho1ij)
    })
  })
# --- Trajectoire temporelle avec les paramètres estimés ---
Z_T2_est <- TemporalSim(nt, r_t, rho1_est, a = rs_est, nu = nu_est, A = A_est)

plot(1:nb_points, Z_T2_est[1:nb_points, 1], type = "l")
points(1:nb_points, Z_T2_est[1:nb_points, 2], type = "l", lty = 2)
points(1:nb_points, Z_T2_est[1:nb_points, 3], type = "l", lty = 3)
```

## Plotting at time 1 precipitation and Temp_max

```{r eval=FALSE}
ns <- nrow(spatial_coordinates)

Z = Z1
for (i in seq_len(ns)){
  for (k in seq_len(p)){
    Z[,i,k] = Z_T2_est[,k]*Z1[,i,k] + Z_T1[,k]
  }
}

# Plot the variable
x_coords = spatial_coordinates[, 1]
y_coords = spatial_coordinates[, 2]

df_v1 = data.frame(
  ZZ1      = Z[1, , 2],
  X        = x_coords,
  Y        = y_coords,
  Variable = paste("V1"),
  Time     =  1
)

df_v2 = data.frame(
  ZZ2      = Z[1, , 3],
  X        = x_coords,
  Y        = y_coords,
  Variable = paste("V3"),
  Time     =  1
)


df_list = list()
df_list[[length(df_list) + 1]] = df_v1
df_list[[length(df_list) + 1]] = df_v2
df = dplyr::bind_rows(df_list)
df = df %>% dplyr::mutate(Time = "Time 1")

time_steps_vec = 1
df_var1 = df %>% filter(Variable == "V1")
df_var2 = df %>% filter(Variable == "V3")

color_scale_fun = function(...) scale_fill_viridis_c(option = "H", ...)
theme_fun = theme_bw

p_1 = ggplot(df_var1, aes(x = X, y = Y, fill = ZZ1)) +
  geom_tile(width = 0.1, height = 0.1) +
  color_scale_fun(name = "Precipitation") +
  #facet_wrap(~ Time, nrow = 1) +
  theme_fun() +
  labs(
    title = paste("Precipitation ; time=1"),
    x = "x coordinate",
    y = "y coordinate",
    fill = NULL
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12),
    strip.text = element_text(size = 10)
  )


p_2 = ggplot(df_var2, aes(x = X, y = Y, fill = ZZ2)) +
  #geom_raster() +
  geom_tile(width = 0.1, height = 0.1) +
  color_scale_fun(name = "Temp_max") +
  #facet_wrap(~ Time, nrow = 1) +
  theme_fun() +
  labs(
    title = paste("Temp_max ; time = 1"),
    x = "x coordinate",
    y = "y coordinate",
    fill = NULL
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12),
    strip.text = element_text(size = 10)
  ) 
pl = p_1 + p_2 + plot_layout(guides = 'collect')
pl
```
