---
title: "Estimating MSTWeatherGen from a Z1 simulation"
description: "Load, inspect, plot, and estimate a non-seasonal MSTWeatherGen model from Z1"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimating MSTWeatherGen from a Z1 simulation}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This vignette shows how to load a simulated three-dimensional array named
`Z1`, inspect its structure, plot one of its variables through time, and fit a
non-seasonal `MSTWeatherGen` model. It uses only the simulated `Z1` dataset and
the 108 spatial coordinates supplied with the package. The original SAFRAN
weather array is not loaded or used.

The expected organization of `Z1` is:

```text
Z1[time, spatial location, variable]
```

For the example considered here, its expected dimensions are 1,461 days, 108
spatial locations, and 3 variables.

## Load MSTWeatherGen

```{r load-package, message=FALSE, warning=FALSE}
library(MSTWeatherGen)
```

## Part 1 :  Read the Z1 simulation data

The code below searches for `Z1_V2.rds` from either the package source tree or
the local project. For an installed package, the recommended location is
`inst/extdata/Z1_V2.rds`.

```{r read-z1}
# Lecture du fichier Z1
fichier_Z1 <- "/home/aboualam/MSTWeatherGen/data/Z1_V2.rds"

# Vérification et lecture
stopifnot(file.exists(fichier_Z1))
Z1 <- readRDS(z1_path)

if (length(dim(Z1)) != 3L) {
  stop("Z1 must be a three-dimensional array: time, location, variable.")
}

dim(Z1)
```

## Part 2 : Define dates, coordinates, and variable names

The simulation covers the period from 1 January 2018 to 31 December 2021.
The coordinates distributed with `MSTWeatherGen` contain the 108 locations
used here.

```{r define-metadata}
dates_Z1 <- seq(
  from = as.Date("2018-01-01"),
  to   = as.Date("2021-12-31"),
  by   = "day"
)

utils::data("coordinates", package = "MSTWeatherGen")

coordinates_Z1 <- coordinates[, c("longitude", "latitude")]

weather_names <- c(
  "Precipitation",
  "Wind",
  "Temp_max"
)

expected_dimensions <- c(
  length(dates_Z1),
  nrow(coordinates_Z1),
  length(weather_names)
)

dimnames(Z1)[[3]] <- weather_names

dim(Z1)
head(dates_Z1)
head(coordinates_Z1)
```


## Plot one variable through time A reverifier.

The following function plots one variable at one spatial location. The
defaults display the third variable (`Temp_max`) at the first location.

```{r plot-function}
plot_Z1_variable <- function(
  Z,
  dates,
  variable = 3,
  position = 1,
  variable_names = dimnames(Z)[[3]]
) {
  if (length(dim(Z)) != 3L) {
    stop("Z must have dimensions: time, location, variable.")
  }

  if (length(dates) != dim(Z)[1]) {
    stop("The number of dates must equal the first dimension of Z.")
  }

  if (variable < 1L || variable > dim(Z)[3]) {
    stop("The requested variable does not exist.")
  }

  if (position < 1L || position > dim(Z)[2]) {
    stop("The requested spatial position does not exist.")
  }

  variable_label <- if (is.null(variable_names)) {
    paste("Variable", variable)
  } else {
    variable_names[variable]
  }

  plot(
    dates,
    Z[, position, variable],
    type = "l",
    lwd = 1,
    col = "steelblue4",
    xlab = "Date",
    ylab = variable_label,
    main = paste(variable_label, "- spatial location", position)
  )
}
```

```{r plot-z1}
plot_Z1_variable(
  Z = Z1,
  dates = dates_Z1,
  variable = 3,
  position = 1
)
```

## Estimate a non-seasonal model

The estimation is performed over the full 2018--2021 period. Consequently,
`by_season` is set to `FALSE` and no `seasons` argument is supplied.

`scale` is also set to `FALSE`: the internal seasonal standardization requires
more than five years of observations, whereas this example contains four
years.

The following computation can take time, so this chunk is not run while the
vignette is built. Remove `eval=FALSE` or run the code interactively to perform
the estimation.

```{r estimate-z1, eval=FALSE}
seasons <- list(
  s1 = list(min_day = 1, max_day = 29, min_month = 12, max_month = 2),
  s2 = list(min_day = 1, max_day = 31, min_month = 3, max_month = 5),
  s3 = list(min_day = 1, max_day = 31, min_month = 6, max_month = 8),
  s4 = list(min_day = 1, max_day = 30, min_month = 9, max_month = 11)
)

swg_Z1 <- MSTWeatherGen_Estim(
  data                = Z1,
  seasons             = seasons,
  dates               = dates_Z1,
  names               = weather_names,
  by_season           = TRUE,
  scale               = FALSE,
  precipitation       = FALSE,
  names_weather_types = weather_names,
  coordinates         = coordinates,
  max_it              = 100,
  tmax                 = 1,
  n1                   = 3,
  n2                   = 3
)
```

If the first variable is a Gaussian field containing negative values rather
than a non-negative precipitation field, rename it and disable the special
precipitation treatment before estimation:

```{r estimate-gaussian-z1, eval=FALSE}
gaussian_names <- c("Variable_1", "Wind", "Temp_max")
dimnames(Z1)[[3]] <- gaussian_names

swg_Z1 <- MSTWeatherGen_Estim(
  data                = Z1,
  dates               = dates_Z1,
  names               = gaussian_names,
  by_season           = FALSE,
  scale               = FALSE,
  precipitation       = FALSE,
  names_weather_types = gaussian_names,
  coordinates         = coordinates_Z1,
  max_it              = 100,
  tmax                 = 1,
  n1                   = 3,
  n2                   = 3
)
```

## Inspect and save the fitted model

After running one of the estimation chunks, inspect and save the resulting
object as follows:

```{r save-estimation, eval=FALSE}
str(swg_Z1, max.level = 2)

saveRDS(
  swg_Z1,
  file = "swg_Z1_estimation.rds"
)
```

## Summary

This workflow:

1. loads only the simulated `Z1` array;
2. associates it with 1,461 dates and 108 coordinates;
3. validates its `time x location x variable` structure;
4. plots a selected variable through time; and
5. estimates one non-seasonal `MSTWeatherGen` model over the complete period.
