---
title: "Estimating MSTWeatherGen from a Z1 simulation"
description: "Load, inspect, plot, and estimate a non-seasonal MSTWeatherGen model from Z1"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimating MSTWeatherGen from a Z1 simulation}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

> Le plan : Introduction, generation de Z1, lancer l'estimation du modèle, comparer les paramètres et valeurs retourné.

## Introduction

The goal of this vignette is to test the estimation phase of MSTWeatherGen and especially the function `estimation_gf` 
in GRF_helper_function which is an internal function.  
Here is how to load the package internal function that is used.

```{r load-package, message=FALSE, warning=FALSE}
devtools::load_all()
# Set the number of cores to use
cores_nb <- setCores(1)
```

## Estimate from know parameters

Here we shows how to create a simulated a latent gaussien random field named `Z` of three-dimensional array that fit a non-seasonal `MSTWeatherGen` model estimation.  
The expected organization of `Z` is:  
`Z[time, spatial location, variable]`

### Generate Z

```{r , eval=FALSE}
library(invgamma)

# Set seed if you want
#set.seed(14347)

# Define spatial and temporal resolutions
spatial_resolution  = 0.4
temporal_resolution = 1 # nb of day

  # Generate spatial grid for simulation
s1_seq = seq(0, 10, by = spatial_resolution)
s2_seq = seq(0, 8, by = spatial_resolution)
grid_spatial = expand.grid(s1 = s1_seq, s2 = s2_seq)
spatial_coordinates = as.matrix(grid_spatial) 
# Define temporal sequence 40 days
t_seq = seq(1, 40, by = temporal_resolution)

generate_Z <- function(t_seq, spatial_coordinates){
  ns = nrow(spatial_coordinates)   
  nt = length(t_seq)  

  p = 3 # Number of variables observed
  L = 3000 # Number of waves

  # Définition des Matern
  rs = c(0.5,1,2) # paramètres aii dans le papier
  nu = c(0.5,1,2) # nuii

  # Define Matern covariance function parameters
  matern_fun1 = function(t_val, s_coord) {
    list(nu=nu[1], r=rs[1])
  }
  matern_fun2 = function(t_val, s_coord) {
    list(nu=nu[2], r=rs[2])
  }
  matern_fun3 = function(t_val, s_coord) {
    list(nu=nu[3], r=rs[3])
  }
  SpectralPars = list(matern_fun1, matern_fun2, matern_fun3)

  # Define covariance structure (between variables) function
  Sigma_fun = function(t_val, s_coord) {
    M = matrix(0.7,ncol=3,nrow=3)
    diag(M) = 1
    return(M)
  }

  # Define anisotropy functions for Isotropic case
  aniso_fun1 = aniso_fun2 = aniso_fun3 = function(t_val, s_coord) {
    diag(2)
  }
  anisotropies = list(aniso_fun1, aniso_fun2, aniso_fun3)

  # Define simulation parameters
  params = list(
    p = p,
    nt = nt,
    t = t_seq,
    A = c(0.2,0.5,0.8), # paramètres A_i
    a = 1/2,   # a=d
    alpha = 1, # b=e, with c=1
    delta = 0, # does not exist in MSTWeatherGen
    b = 1, # non-separability - does not exist in MSTWeatherGen
    anisotropies = anisotropies,
    SpectralPars = SpectralPars,
    Sigma_fun    = Sigma_fun,
    rs = rs,
    nu = nu)

  # Run spatio-temporal simulation 
  Z1 = SimulateParsimNS(
    L                  = L,
    ns                 = ns,
    nt                 = nt,
    p                  = p,
    params             = params,
    spatial_coordinates = spatial_coordinates,
    SpectralDensity = MaternSpectralDensity3D,
    parallelize        = F, 
    batch_size = 100,
    n_cores = 4
  )

  dimnames(Z1)[[3]] <- c("v1","v2","v3")

  r_t = c(0.2,0.4,0.6) # r¹ii et r²ii
  rho  = Sigma_fun(1,1)
  Z_T1 = TemporalSim(nt,r_t,rho,a=rep(1,p),nu=rep(1,p),A=rep(0,p))
  Z_T2 = TemporalSim(nt,r_t,rho,a=rs,nu=nu,A=c(0.2,0.5,0.8))

  Z = Z1
  for (i in seq_len(ns)) {
    for (k in seq_len(p)){
      Z[,i,k] = 0.707*(Z_T2[,k]*Z1[,i,k] + Z_T1[,k])
    }
  }

  #plot(1:40,Z_T1[,1],type="l")
  #points(1:40,Z_T1[,2],type="l",lty=2)
  #points(1:40,Z_T1[,3],type="l",lty=3)

  #plot(1:40,Z_T2[,1],type="l")
  #points(1:40,Z_T2[,2],type="l",lty=2)
  #points(1:40,Z_T2[,3],type="l",lty=3)

  #plot(1:40,Z[,1,1],type="l")
  #points(1:40,Z[,1,2],type="l",lty=2)
  #points(1:40,Z[,1,3],type="l",lty=3)

  return(Z)
}

```

### Check Z

Plot two variables at a time T
```{r eval=FALSE}
library(dplyr)

Z = generate_Z(t_seq, spatial_coordinates)

T = 1
variables = c("v1","v3")
x_coords = spatial_coordinates[, 1]
y_coords = spatial_coordinates[, 2]

df_v1 = data.frame(
  ZZ1      = Z[T, , 2],
  X        = x_coords,
  Y        = y_coords,
  Variable = paste(variables[1]),
  Time     =  1
)

df_v2 = data.frame(
  ZZ2      = Z[T, , 3],
  X        = x_coords,
  Y        = y_coords,
  Variable = paste(variables[2]),
  Time     =  T
)

color_scale_fun = function(...) scale_fill_viridis_c(option = "H", ...)
theme_fun = theme_bw

p_1 = ggplot(df_v1, aes(x = X, y = Y, fill = ZZ1)) +
  geom_raster() +
  color_scale_fun() +
  #facet_wrap(~ Time, nrow = 1) +
  theme_fun() +
  labs(
    title = paste("Variable=", variables[1] ,"; time=",T),
    x = "x coordinate",
    y = "y coordinate",
    fill = NULL
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12),
    strip.text = element_text(size = 10)
  )


p_2 = ggplot(df_v2, aes(x = X, y = Y, fill = ZZ2)) +
  geom_raster() +
  color_scale_fun() +
  #facet_wrap(~ Time, nrow = 1) +
  theme_fun() +
  labs(
    title = paste("variable=",variables[2],"; time = ",T),
    x = "x coordinate",
    y = "y coordinate",
    fill = NULL
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12),
    strip.text = element_text(size = 10)
  ) 
pl = p_1 + p_2 + plot_layout(guides = 'collect')
pl

```


### Run estimation  

This is how to run Z simulation and estimation.

#### Part 1 : Read the already saved Z simulation data 

```{r eval=FALSE}
# to check, Z is simulated later
data("Z", package = "MSTWeatherGen")

if (length(dim(Z)) != 3L) {
  stop("Z must be a three-dimensional array: time, location, variable.")
}
```

#### Part 2 : Define estimation_gf parameters

```{r eval=FALSE}
tmax <- 3 # nb of days of lag
wt_id = (tmax+1):(length(t_seq)) # one weather type , all indexes starting at tmax+1 match t_seq and dates

max_it <- 100 # Maximum number of iteration for optimisation process

dates = seq(as.Date('2018-01-01'),as.Date('2018-01-01')+(length(t_seq)-1),by = 'day')
names = c('v1','v2','v3')
n1 = 10
n2 = 10
coordinates = as.data.frame(spatial_coordinates)
names(coordinates) = c('longitude','latitude')

rho1_init <- matrix(0.1, nrow = length(names), ncol = length(names))
diag(rho1_init) <- 0.5
dimnames(Z)[[3]] = names

cr <- sapply(names, function(v1) {
  sapply(names, function(v2) {
    mean(sapply(1:dim(Z)[2], function(j)
      cor(Z[,j,v1], Z[,j,v2], use="complete.obs")), na.rm=TRUE)
  })
})
cr <- matrix(cr, nrow=length(names), ncol=length(names))
colnames(cr) <- rownames(cr) <- names
```

#### Part 3 Simulations. 

```{r eval=FALSE}
Sys.setenv(MONITOR = "TRUE")
nb_estimation <- 1
estim_results <- list()

for( it in 1:nb_estimation) {

  #set.seed(14357+it)
  Z <- generate_Z(t_seq, spatial_coordinates)

  estim_results[[it]] <- estimation_gf(
      data             = Z,
      wt_id            = wt_id,
      max_it           = max_it,
      dates            = dates,
      tmax             = tmax,
      names            = names,
      coordinates      = coordinates,
      n1               = n1,
      n2               = n2,
      rho1             = rho1_init,
      cr               = cr,
      threshold_precip = rep(0,dim(Z)[2])
  )
}

library(ggplot2)

# Lire le fichier
spatial       <- read.csv("monitoring_aii_nuii_spatial.txt", sep=";")
spatial_iters <- spatial[grepl("iter_", spatial$label), ]
spatial_iters$iter <- seq_len(nrow(spatial_iters))

# Colonnes aii et nuii
aii_cols  <- grep("^aii_",  colnames(spatial_iters), value=TRUE)
nuii_cols <- grep("^nuii_", colnames(spatial_iters), value=TRUE)

# Format long pour aii
df_aii <- data.frame(
  iter     = rep(spatial_iters$iter, length(aii_cols)),
  variable = rep(gsub("aii_", "", aii_cols), each=nrow(spatial_iters)),
  value    = unlist(spatial_iters[, aii_cols])
)

# Format long pour nuii
df_nuii <- data.frame(
  iter     = rep(spatial_iters$iter, length(nuii_cols)),
  variable = rep(gsub("nuii_", "", nuii_cols), each=nrow(spatial_iters)),
  value    = unlist(spatial_iters[, nuii_cols])
)

# Filtrer jusqu'à l'itération 26
df_aii_26  <- df_aii[df_aii$iter  <= 26, ]
df_nuii_26 <- df_nuii[df_nuii$iter <= 26, ]

# Plot aii jusqu'à iter 26
# Plot aii jusqu'à iter 26
p_aii_26 <- ggplot(df_aii_26, aes(x=iter, y=value, color=variable)) +
  geom_line(alpha=0.7) +
  theme_minimal() +
  labs(title="Evolution de aii — iter 1 à 26", x="Itération", y="aii", color="Variable") +
  theme(plot.title=element_text(hjust=0.5))

# Plot nuii jusqu'à iter 26
p_nuii_26 <- ggplot(df_nuii_26, aes(x=iter, y=value, color=variable)) +
  geom_line(alpha=0.7) +
  theme_minimal() +
  labs(title="Evolution de nuii — iter 1 à 26", x="Itération", y="nuii", color="Variable") +
  theme(plot.title=element_text(hjust=0.5))

print(p_aii_26)
print(p_nuii_26)

print(p_aii_26)
print(p_nuii_26)

```

### Comparing parameters values

#### Parameters values from estimation

```{r eval=FALSE}
# get all par_all values from all estimation
all_estimations <- do.call(rbind,lapply(estim_results, function(est){return(est$par_all)}))

all_estimations_stats <- data.frame(
  moyenne = apply(all_estimations, 2, mean, na.rm = TRUE),
  variance = apply(all_estimations, 2, var, na.rm = TRUE),
  ecart_type = apply(all_estimations, 2, sd, na.rm = TRUE),
  min = apply(all_estimations, 2, min, na.rm = TRUE),
  max = apply(all_estimations, 2, max, na.rm = TRUE),
  mediane = apply(all_estimations, 2, median, na.rm = TRUE)
)

all_estimations_stats

```

```{r eval=FALSE}
cols_num <- c("a","b","c","d","e","Ai","Aj","aii","ajj",
              "nuii","nujj","r1ii","r1jj","r2ii","r2jj",
              "rho1ij","rho2ij")

# Mean
parm_all_runs <- do.call(rbind, lapply(1:nb_estimation, function(k) estim_results[[k]]$parm))
parm_mean <- aggregate(parm_all_runs[, cols_num], 
            by = list(v1=parm_all_runs$v1, v2=parm_all_runs$v2), 
            FUN = mean, na.rm = TRUE)

# estimated parameters mean values
parm_mean
```

#### Comparing real parameters and estimated parameters values

On met les parametres initiaux dans un dataframe
```{r eval=FALSE}
# --- Table des vraies valeurs, structurée comme parm_mean ---

# Toutes les paires (v1, v2), diagonale + hors-diagonale, comme dans mean_global
var_pairs <- expand.grid(v1 = names, v2 = names, stringsAsFactors = FALSE)
var_pairs <- var_pairs[as.character(var_pairs$v1) <= as.character(var_pairs$v2), ]

# Indices correspondants dans rs / nu / A (dans l'ordre de `names`)
i <- match(var_pairs$v1, names)
j <- match(var_pairs$v2, names)

rho_1  = Sigma_fun(1,1)

r_t = c(0.2,0.4,0.6)

true_params <- data.frame(
  v1     = var_pairs$v1,
  v2     = var_pairs$v2,
  a      = params$a,      
  b      = params$alpha,   
  c      = 1,              
  d      = params$a,       
  e      = params$alpha,   
  Ai     = params$A[i],
  Aj     = params$A[j],
  aii    = params$rs[i],
  ajj    = params$rs[j],
  nuii   = params$nu[i],
  nujj   = params$nu[j],
  rho1ij = mapply(function(a, b) rho_1[a, b], i, j),
  r2ii   = r_t[i],
  r2jj   = r_t[j],
  r1ii   = r_t[i],
  r1jj   = r_t[j],
  rho2ij = mapply(function(a, b) rho_1[a, b], i, j)
)

# diff estimated vs true_params
parm_mean[,-c(1,2)] - true_params[,-c(1,2)]
```
